
---
title: "Data Wrangling with  R"
subtitle: "IBM 320 Section 02"
author: "ion Peoples"
date: "2026-03-04"
format:
    html:
      theme: cosmo
      toc: true
      toc-depth: 3
      tocc-location: left
      embed-resources: 
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
execute:
  freeze: auto
---

## learning outcome

:::: callout-tip
-Describe the concept of Data Wrangling.
-Describe how Tibbles are different from data frames.
-Explain how to convert wide or long data to "Tidy" data.
-Explain how to use grouped mutates and filter together.
-Be familiar with major dplyr functions for transforming data.
-Create a new variable with mutate() and case_when().
-Use the pipe operator to shape the data to prepare for analysis and visualization.

too use multi-cursor editing feature, do the following steps:

::: callout-tip
To use mouse multi-cursor editing feature, do the following steps: 1.
Select the text you want to edit.
2.
Press `shift + alt + i` to place a cursor at the end of each selected line.
3.
You can now edit all the selected text simultaneously.
:::

# Introduction to Data Wrangling

##loading packages **Note** To add a code chunk, you can use the keyboard shortcut 'Ctrl+ Alt + I'.

```{r}
#install.packages("tidyverse")
#install.packages("nycflights13")

library(tidyverse)
library(nycflights13)

```

## Tibbles vs Data Frames

```{r}
class(mtcars)
cars.tib <- mtcars |>
  rownames_to_column() |> 
  as_tibble() |> 
  rename(model = rowname)
  
cars.tib
class(cars.tib)
typeof(cars.tib)
  
```